prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-21 inst/MyArticle/paper.Rmd
reviewed text. Still references to supplement are missing.

@@

The probabilistic dropout analysis implemented in  `r BiocStyle::Biocpkg("proDA")`, as well as the hurdle model implemented in `r BiocStyle::Biocpkg("msqrob2")`, produce inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A). Because of the robustness of the dropout and hurdle model to missing observations we obtain difference estimates for all proteins and contrasts (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated differences of the probabilistic dropout model are unbiased and show high diagnostic accuracy (Figure \@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation (`prot_med_lm_moderated`). Therefore, we are planning to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

The R-package `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("prolfqua")` model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` does not estimate group differences for more proteins and does not achieve a higher $pAUC$ score than `r BiocStyle::Githubpkg("fgcz/prolfqua")`. Furthermore, Figure \@ref(fig:FDRfdp) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low $FDR$.

We focused our benchmark on comparing the statistical modeling methods using three different scores, while we fixed the pre-processing steps. However, there are other equally or even more important parameters of a protein quantification pipeline [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [@Grossmann2010]. For instance, the original `r BiocStyle::Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is significantly lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 73%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast], identifying proteins and generating the quantification values can also significantly contribute to the performance of the entire pipeline, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis.

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is an feature-rich, object oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure 2 and Table 3). The contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework is flexible enough to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables you to call various methods and quickly select the best differential expression analysis algorithm for your problem. It also can generate conclusive reports and can be used to benchmark MS software and statistical methods. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the $FDR$ is an unbiased estimate of the $FDP$. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis results obtained with prolfqua are comparable to or better than those of other differential expression analysis tools.

>>>>

\begin{quote}
To paraphrase provocatively, `machine learning is statistics minus any checking of models and assumptions'.
\begin{flushright}
-- Brian D. Ripley, useR! 2004, Vienna.
\end{flushright}
\end{quote}

@@

The R-package `r BiocStyle::Biocpkg("proDA")`, `r BiocStyle::Biocpkg("msqrob2")` and `r BiocStyle::Biocpkg("prolfqua")` (see Methods) model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` estimates group differences for less proteins and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:FDRfdp) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low $FDR$.

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:FDRfdp) Panel A when using the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the other methods (Figure \@ref(fig:FDRfdp) Panel B). All three models perform similarly well when benchmarked against the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (supplement). We tested if the $pAUC_{10}$ differences for the three methods were statistically significant, and none of them were at a significance level of $0.4$ (see supplement). Furthermore, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:FDRfdp) Panel C) are comparable for all three methods.

Since, the API of `r BiocStyle::Biocpkg("proDA")` and the produced outputs are similar to those of prolfqua, we are planning to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)), to make it easy for application developers to use both.

We focused our benchmark on comparing the statistical modeling methods using three different scores while we fixed the pre-processing steps. However, there are other equally or even more important parameters of a protein quantification pipeline [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer protein intensities from peptide intensities is an additional important factor [@Grossmann2010]. For instance, the original `r BiocStyle::Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is significantly lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptide abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 75%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast], identifying proteins and generating the quantification values, significantly contributes to the entire pipeline's performance, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the supplement we show the benchmarking results for `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Githubpkg("fgcz/prolfqua")` when the IonStar dataset is prepossessed with the MSFragger/IonQuant [@yu2020fast] quantification pipeline (v15). Although the number of proteins identified with two peptides is slightly lower than with MaxQuant (3929 vs. 4046), the $pAUC_{10}$ is significantly higher ($81%$ vs. $72%$).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is a feature-rich, object-oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure 2 and Table 3). Furthermore, the contrast specification is explicit and consistent for all models and allows for testing interactions. Furthermore, the modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework is flexible enough to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables the implementation of applications where the user can select an alternative normalization method, protein abundance estimation method, or differential expression analysis algorithm. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the $FDR$ is an unbiased estimate of the $FDP$. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; and last but not least vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis results obtained with prolfqua are comparable to those of other differential expression analysis tools.
